---
title: "Homework 5 - ST 541 PCS "
author: "Elena Volpi"
date: "11/15/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

#Problem 2.1 
The following data are an i.i.d sample from a Cauchy($\theta$,1) distribution:
```{r}
cauchy_sample<- c(1.77,-.23,2.76,3.8,3.47,56.75, -1.34,4.24,-2.44,3.29,3.71,-2.4,4.53,-0.07,-1.05,-13.87,-2.53,-1.75,.27,43.21)
cauchy_sample
``` 


a) Graph the log likelihood function. Find the MLE for $\theta$ using the Newton-Raphson method. Try all of the following starting points: -11, -1, 0, 1.5,4,4.7,7,8, and 38. Discuss your results. Is the mean of the data a good starting point?

We find the log likelihood function 

```{r}
#Log likelihood plot 
theta_val <- seq(-50,50,.01)
log_likelihood <-  function(x_0, y= cauchy_sample){
  sum(dcauchy(y, location = x_0, scale = 1, log = TRUE))
}

y = map_dbl(theta_val,log_likelihood)
plot(y ~ theta_val, ylab = "Log likelihood", xlab = expression(theta), main = "Log likelihood for Cauchy with scale 1")
```

We find the MLE 

```{r}


#First derivative 
llhood_cauchy_derivative <- function(x_0, y = cauchy_sample){
  sum(2 * (y - x_0)/(1 + (y - x_0)^2))
}

#Second Derivative 
llhood_cauchy_derivative2 <- function(x_0, y = cauchy_sample){
  n <- length(y)
  sum((-2 + 2 * (y - x_0)^2)/(1 + (y - x_0)^2)^2)
}

newton <- function(x,theta,m=10000,tol=0.00001){
  iter <-0
  old_theta <- theta
  theta <- old_theta - (llhood_cauchy_derivative(x=x,theta=old_theta)/llhood_cauchy_derivative2(x=x,theta=old_theta))
  while(abs(theta-old_theta)>tol){
    iter <- iter +1
    if (iter > m) 
      stop("No solution found")
    old_theta <- theta 
    theta <- old_theta - (llhood_cauchy_derivative(x,old_theta)/llhood_cauchy_derivative2(x=x,theta=old_theta))
  }
  
  return(list(theta,iter))
}

sp <- c(-11,-1,0,1.5, 4, 4.7, 7, 8, 38)
MLEs <- map(sp,~newton(x=cauchy_sample, theta= .))



```


```{r}
llhood_cauchy_derivative <- function(x_0, y = cauchy_sample){
  sum(2 * (y - x_0)/(1 + (y - x_0)^2))
}

#Second Derivative 
llhood_cauchy_derivative2 <- function(x_0, y = cauchy_sample){
  n <- length(y)
  sum((-2 + 2 * (y - x_0)^2)/(1 + (y - x_0)^2)^2)
}

newton_print <- function (f, fp, x, tol = 0.00001, m = 100) 
{
    iter <- 0
    oldx <- x
    x <- oldx + 10 * tol
    while (abs(x - oldx) > tol) {
        iter <- iter + 1
        if (iter > m) 
            stop("No solution found")
        oldx <- x
        x <- x - f(x)/fp(x)
    }
    info <<- list(x,iter)
    return(info)
}

sp <- c(-1,0,1.5, 4, 4.7, 7,38) #the ones that provide solutions
MLEs <- map(sp,~newton_print(f = llhood_cauchy_derivative, fp =llhood_cauchy_derivative2, x=.))


newton_print(llhood_cauchy_derivative,fp =llhood_cauchy_derivative2, x = mean(cauchy_sample))


```

There was no solution to the starting point values -11 and 8. Starting point values of -1, 0, 1.5, 4, 4.7, 7, 38 provides MLE values of -0.19, -0.19, 1.71,2.81, -0.19, 41.04, 42.79, respectively. The mean provided a MLE from the newton method of 54.88. Our best estimates were -1, 0, 4.7 since they were closest to theta and provided the fewest iterations. The mean was not a good starting point as it took 9 iterations to obtain the MLE. 

b) Apply the bisection method with starting points -1 and 1. Use additional runs to illustrate manners in which the bisection method may fail to find the global maximum.

```{r}


bisection<- function (f,x, a, b, tol = 0.001, m = 100) {
    iter <- 0
    f.a <- f(x,a)
    f.b <- f(x,b)
    while (abs(b - a) > tol) {
        iter <- iter + 1
        if (iter > m) {
            break
        }
        xmid <- (a + b)/2
        ymid <- f(x,xmid)
        if (f.a * ymid > 0) {
            a <- xmid
            f.a <- ymid
        }
        else {
            b <- xmid
            f.b <- ymid
        }
    }
    root <- (a + b)/2
    return(list(root, iter))
}
bisection(llhood_cauchy_derivative, x = cauchy_sample, a=-1, b= 1)
bisection(llhood_cauchy_derivative, x = cauchy_sample, a=-.5, b= 5)
bisection(llhood_cauchy_derivative, x = cauchy_sample, a=.5, b= -5)
bisection(llhood_cauchy_derivative, x = cauchy_sample, a=-.5, b= -.5)
```
It takes 11 iterations to obtain the MLE. It seems to use endpoints if it cannot find a maximum. 

c) 
```{r}
fixed_point_print <- function (f, x, tol = 0.001, m = 1000) 
{
    iter <- 0
    oldx <- x
    x <- oldx + 10 * tol
    while (abs(x - oldx) > tol) {
        iter <- iter + 1
        if (iter > m) 
            stop("No solution found")
        oldx <- x
        x <- f(x)
    }
    return(x)
}
```

